Benchmarking R packages for calculation of persistent homology

### 2020-12-28 tdabench-paper/tdabench.Rmd
updated discussion for added memory benchmarks

However, without sufficient computational resources, alpha complexes were not usable for point cloud dimensions greater than 3.
If a point cloud has more than 3 dimensions, then it could undergo pre-processing with dimension reduction before using alpha complexes.
Note, it is possible an algorithm will eventually be developed to enable alpha complex computation of higher dimensional data. 
However, if data dimension cannot be reduced without significant information loss, then Vietoris-Rips complexes should be used.
It should be stated that if a point cloud is compatible with both complexes, both analyses should be performed as there may be a variation the persistent homology matrix. 
This is because alpha complexes satisfy the Nerve Theorem \citep{alpha-complex}, which implies that they are topologically equivalent to the true underlying topology of the dataset; in contrast, Vietoris-Rips complexes only approximate the underlying topology \citep{Rips-Complex}.
Among the tested libraries, Ripser (wrapped by \CRANpkg{TDAstats}) has the fastest calculation time. 
GUDHI and Dionysus (wrapped by \CRANpkg{TDA}) significantly fall behind as feature dimension and number of points increase. 
GUDHI and Dionysus, however, do have the option of returning the boundary matrix, which Ripser does not store for the sake of optimization.
While we did not measure Ripser's boundary matrix object size, Ripser's C++ code (outside its implementation in R) demonstrates memory advantages \citep{Ripser}

On average, Ripser wrapped by \CRANpkg{TDAstats} computed persistent homology of a Vietoris-Rips complex faster and with less memory than either GUDHI or Dionysus wrapped by \CRANpkg{TDA}.
Thus, when efficiency is critical, useRs would likely find \CRANpkg{TDAstats} the appropriate library.
However, \CRANpkg{TDA} contains an entire library of features not available in \CRANpkg{TDAstats}.
Specifically, \CRANpkg{TDA} allows kNN density estimation, kernel density estimation, density-based clustering, and dendrogram visualization, among other useful functionality.
When computational resources are plenty, when point clouds are small and low-dimensional, or when the aforementioned functionality is required, \CRANpkg{TDA} will likely be more appropriate than \CRANpkg{TDAstats}.
Both packages are hosted on CRAN.

>>>>

However, without sufficient computational resources, alpha complexes were not usable for point cloud dimensions greater than 3.
If a point cloud has more than 3 dimensions, then it could undergo pre-processing with dimension reduction before using alpha complexes.
Note, it is possible an algorithm will eventually be developed to enable alpha complex computation of higher dimensional data. 
However, if data dimension cannot be reduced without significant information loss, then Vietoris-Rips complexes should be used.
It should be stated that if a point cloud is compatible with both complexes, both analyses should be performed as there may be a variation the persistent homology matrix. 
This is because alpha complexes satisfy the Nerve Theorem \citep{alpha-complex}, which implies that they are topologically equivalent to the true underlying topology of the dataset; in contrast, Vietoris-Rips complexes only approximate the underlying topology \citep{Rips-Complex}.
Among the tested Vietoris-Rips engines, Ripser (wrapped by \CRANpkg{TDAstats}) has the fastest calculation time. 
GUDHI and Dionysus (wrapped by \CRANpkg{TDA}) significantly fall behind as feature dimension and number of points increase.

On average, Ripser computed persistent homology of a Vietoris-Rips complex with less memory than either GUDHI or Dionysus.
Thus, when efficiency is critical, useRs would likely find \CRANpkg{TDAstats} the appropriate library.
However, \CRANpkg{TDA} contains an entire library of features not available in \CRANpkg{TDAstats}.
Specifically, \CRANpkg{TDA} allows kNN density estimation, kernel density estimation, density-based clustering, and dendrogram visualization, among other useful functionality.
When computational resources are plenty, when point clouds are small and low-dimensional, or when the aforementioned functionality is required, \CRANpkg{TDA} will likely be more appropriate than \CRANpkg{TDAstats}.
Both packages are hosted on CRAN.
